---
title: "Research Worldwide Bureaucracy Indicators to help implement efficient public policies"
subtitle: "INFO 526 - Summer 2024 - Final Project"
author: "Mohammad Rahat Helal"
title-slide-attributes:
  data-background-image: None
  data-background-size: stretch
  data-background-opacity: "0.7"
  data-slide-number: none
format:
  revealjs:
    theme:  "data/customtheming.scss"
editor: visual
execute:
  echo: true
---

```{r}
#| label: load-packages
#| message: false
#| output: false
#| echo: false

if(!require(pacman))
  install.packages("pacman")

pacman::p_load(
  here,
  tidyverse,
  ggrepel,
  devtools,
  tidytuesdayR,
  dplyr,
  readr,
  janitor,
  ggplot2,
  showtext,
  gridExtra,
  sf,
  rnaturalearth,
  rnaturalearthdata,
  skimr
  )

options(width = 65)

knitr::opts_chunk$set(
  fig.asp = 0.618,
  fig.align = "center",
  fig.retina = 1,
  fig.showtext = T
)

```

```{r}
#| label: setup
#| include: false
#| message: false
#| output: false
#| echo: false

# Plot theme
ggplot2::theme_set(ggplot2::theme_minimal(base_size = 11))
# 
# For better figure resolution
knitr::opts_chunk$set(
  fig.retina = 3,
  dpi = 300,
  fig.width = 6,
  fig.asp = 0.618
  )

```

## Introduction & Motivation

-   Selected datasets from Worldwide Bureaucracy Indicators (WWBI)
-   Contain wages, income and employment information in public and private sector globally
-   Utilized to create insights into current outcomes of public policies
-   Understand the trends across geographies, and focus on a particular country
-   How to improve or suggest improvements in country-specific needs and situations

## Dataset - Worldwide Bureaucracy Indicators {.smaller}

```{r}
#| label: load-data
#| message: false
#| echo: false
#| output: false

# Suppress all output during the data loading
invisible(capture.output({
  tuesdata <- tidytuesdayR::tt_load('2024-04-30')
}))

# create separate dataframes for each parts - data, series, and country
wwbi_data <- tuesdata$wwbi_data
wwbi_series <- tuesdata$wwbi_series
wwbi_country <- tuesdata$wwbi_country

```

wwbi_data, wwbi_series, and wwbi_country

```{r}
#| label: show-data
#| message: false
#| echo: false

options(width = 200)  # You can increase this number to a value that suits your needs

glimpse(wwbi_data)
glimpse(wwbi_series)
glimpse(wwbi_country)


```

## Questions

The two questions I want to answer:

### Question 1:

Compare the different regions globally, research how has public sector employment changed over time? Pick up one country in a region and research if there are underlying factors for such changes?

### Question 2:

In context of public administration in different geographies, what is the relationship between income group and public administration efficiency?

## Visualization - 1 (Question 1a)

```{r}
#| label: Visualization-1
#| message: false
#| echo: false

# replicate base data into three dataframes to analyze question 1
country_df <- wwbi_country 
data_df <- wwbi_data 
series_df <- wwbi_series 

# merge the three data frames on indicator code and country code 
merged_df <- merge(data_df, series_df, by = "indicator_code")
merged_df <- merge(merged_df, country_df, by = "country_code")

# Public sector employment
ps_empl_indicator <- merged_df %>%
  filter(str_detect(str_to_lower(indicator_name), "public sector employment"))

distinct_ps_empl_ind <- ps_empl_indicator %>%
  distinct(indicator_code, indicator_name)

grouped_ps_empl_ind <- ps_empl_indicator %>%
  group_by(indicator_code, indicator_name) %>%
  summarise(record_count = n(), .groups = "drop") %>% 
  arrange(desc(record_count))

# Filter for the public sector employment indicator (share of total employment)
employment_data_tot <- merged_df %>%
  filter(indicator_code == "BI.EMP.TOTL.PB.ZS")

# Filter for the public sector employment indicator  (share of total and paid employment)  
empl_data_tot_paid <- merged_df %>%
  filter(indicator_code %in% c("BI.EMP.TOTL.PB.ZS", "BI.EMP.PWRK.PB.ZS", "BI.EMP.FRML.PB.ZS", "BI.EMP.TOTL.PB.UR.ZS", "BI.EMP.TOTL.PB.RU.ZS"))

# Aggregate data by year and region - for All geogrpahies
aggregated_data <- employment_data_tot %>%
  group_by(year, region) %>%
  summarise(mean_value = mean(value, na.rm = TRUE), .groups = "drop")

# Create plot_b with increased width
plot_b <- ggplot(aggregated_data, aes(x = year, y = mean_value, color = region)) +
  geom_line() +
  scale_x_continuous(breaks = unique(aggregated_data$year)) +
  geom_point() +
  labs(title = "Public Sector Employment as a Share of Total Employment\nOver Time by Region - All Geographies",
       x = "Year",
       y = "Public Sector Employment Share (%)",
       color = "Region") +
  theme_minimal() +
  theme(text = element_text(size = 8),
        axis.text.y = element_text(size = 9),
        legend.position = "bottom",
        plot.margin = margin(10, 10, 10, 10),
        legend.key.size = unit(0.7, "lines"),   
        legend.spacing.x = unit(0.5, "cm")    
        )  

print(plot_b)

# Display the plots side by side with adjusted widths
# grid.arrange(plot_b, plot_a, ncol = 2, widths = c(0.55, 0.45))

```

## Visualization - 2 (Question 1b)

```{r}
#| label: Visualization-4
#| message: false
#| echo: false

# Filter for North America region and the two countries
north_america_data <- empl_data_tot_paid %>%
  filter(region == "North America")

# named vector for five indicators
indicator_labels <- c(
  "BI.EMP.TOTL.PB.ZS" = "Share of total employment",
  "BI.EMP.PWRK.PB.ZS" = "Share of paid employment",
  "BI.EMP.FRML.PB.ZS" = "Share of formal employment",
  "BI.EMP.TOTL.PB.UR.ZS" = "Share of total employment by urban location",
  "BI.EMP.TOTL.PB.RU.ZS" = "Share of total employment by rural location"
)

# Add a new column for descriptive labels
north_america_data <- north_america_data %>%
  mutate(indicator_label = factor(indicator_code, levels = names(indicator_labels), labels = indicator_labels))

# Aggregate data by year, country, and indicator_label
aggregated_data_na <- north_america_data %>%
  group_by(year, country_code, indicator_label) %>%
  summarise(mean_value = mean(value, na.rm = TRUE), .groups = "drop")

# Create the grouped bar plot
plot_na <- ggplot(aggregated_data_na, aes(x = year, y = mean_value, fill = country_code)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  facet_wrap(~ indicator_label, scales = "free_y") +
  scale_x_continuous(breaks = unique(aggregated_data_na$year)) +
  labs(title = "Comparison of Public Sector Employment Indicators between\nCanada and United States of America\nOver the Years",
       x = "Year",
       y = "Public Sector Employment Share (%)",
       fill = "Country") +
  theme_minimal() +
  theme(text = element_text(size = 8),
        axis.text.y = element_text(size = 9),
        legend.position = "bottom",
        legend.title = element_blank(),
        legend.text = element_text(size = 8),
        legend.spacing.y = unit(0.5, 'lines'),
        plot.margin = margin(10, 10, 10, 10))

print(plot_na)

```

## Visualization - 3 (Question 2)

```{r}
#| label: Question-2
#| message: false
#| echo: false

country_df_2 <- wwbi_country 
data_df_2 <- wwbi_data 
series_df_2 <- wwbi_series 

# country_df_2 <- country_df_2 %>%
#   clean_names() %>%
#   remove_empty("cols")

# efficiency_indicators <- series_df_2 %>%
#   filter(str_detect(str_to_lower(indicator_name), "public administration efficiency"))

# Check available indicators related to public administration or efficiency
relevant_indicators <- series_df_2 %>%
  filter(str_detect(str_to_lower(indicator_name), "administration|government|efficiency"))

# View relevant indicators
# print(relevant_indicators)

# Filter data_df for these indicators
efficiency_data <- data_df_2 %>%
  filter(indicator_code %in% relevant_indicators$indicator_code)

# Merge data_df with country_df to get income group information
merged_df_2 <- merge(efficiency_data, country_df_2, by = "country_code")

# Aggregate and visualize the data
efficiency_aggregated_2 <- merged_df_2 %>%
  group_by(income_group, year) %>%
  summarise(mean_efficiency = mean(value, na.rm = TRUE), .groups = "drop")

# Plot the data as a facet grid with line graphs
ggplot(efficiency_aggregated_2, aes(x = year, y = mean_efficiency, group = income_group, color = income_group)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ income_group, scales = "free_y") +
  labs(title = "Relationship Between Income Group and Public Administration Efficiency Over Years",
       x = "Year",
       y = "Public Administration Efficiency") +
  theme_minimal() +
  theme(
    text = element_text(size = 8),
    axis.text.x = element_text(hjust = 1),
    legend.position = "none",
    plot.margin = margin(10, 10, 10, 10)
  )

```

## Conclusions and Future Works

-   Researched and plott the trend of public sector employment over the years for all geographies
-   Focus on several indicators between Canada and USA over the years
-   Establish a relationship between different income groups and public administration efficiency over the years for all regions
-   Research the disparities in income groups by gender, wage premium, and other indicators
-   Suggest practices to bridge the gaps between similar economies
